---
output:
  pdf_document: default
  html_document: default
---
# Cleaning Data in R
1. Exploring raw data
2. Tidying data
3. Preparing data for analysis

## Exploring data
```{r}
# pointing to a function in another .R file
source('R_repo/funcs.R')

# reading in a .rds file which contains weather data
weather <- readRDS(file=f('weather.rds'))

head(weather)
```
### Understanding the structure of the data
```{r}
# view the class
class(weather)
```
```{r}
# view data dimensions
D <- dim(weather)
D
paste('DF has', D[1], 'rows/observations and', D[2], 'columns/variables')
```
```{r}
# column names of a DF
names(weather)
```
```{r}
# structure of the DF
str(weather[, 1:5])
```
dplyr allows another alternative to view structure. It gives as much preview of each column as possible.
```{r}
library(dplyr)
glimpse(weather[, 1:5])
```
```{r}
# summary
summary(weather[, 1:5])
```
### Visualizing data
```{r}
# importing the BMI data
bmi <- read.csv(file=f('bmi_clean.csv'), header=TRUE)
dim(bmi)
```
```{r}
# Histogram of BMIs from 2008
hist(bmi$Y2008)

# Scatter plot comparing BMIs from 1980 to those from 2008
plot(x=bmi$Y1980, y=bmi$Y2008)
```
## Tiyding data
## tidyr package
The most important function in tidyr is gather(). It should be used when you have columns that are not variables and you want to collapse them into key-value pairs.
```{r}
# view first 5 rows of bmi
head(bmi, n=5)
```

```{r}
library(tidyr)

# Apply gather() to bmi and save the result as bmi_long
bmi_long <- gather(data=bmi, key=year, value=bmi_val, -Country)

# View the first 5 rows of the result
head(bmi_long, n=5)
```
The opposite of gather() is spread(), which takes key-values pairs and spreads them across multiple columns. This is useful when values in a column should actually be column names (i.e. variables). It can also make data more compact and easier to read.
```{r}
bmi_wide <- spread(data = bmi_long, key = year, value = bmi_val)
head(bmi_wide, n=5)
```
### Separating columns
The separate() function allows you to separate one column into multiple columns. Unless you tell it otherwise, it will attempt to separate on any character that is not a letter or number. You can also specify a specific separator using the sep argument.

```{r}
# create a sample DF
bmi_cc <- bmi_long[1:5, ]
isos <- c("/AF", "/AL", "/DZ", "/AD", "/AO")
bmi_cc$Country <- paste(bmi_cc$Country, isos, sep='')
```
```{r}
names(bmi_cc)[1] <- "Country_ISO"
bmi_cc
```
```{r}
# Apply separate() to bmi_cc
bmi_cc_clean <- separate(bmi_cc, col = Country_ISO, into = c("Country", "ISO"), sep = "/")

# Print the head of the result
head(bmi_cc_clean)
```
### Uniting columns
The opposite of separate() is unite(), which takes multiple columns and pastes them together. By default, the contents of the columns will be separated by underscores in the new column, but this behavior can be altered via the sep argument.
```{r}
# Apply unite() to bmi_cc_clean
bmi_cc <- unite(bmi_cc_clean, col='Country_ISO', 'Country', 'ISO', sep = "-")

# View the head of the result
head(bmi_cc)
```

## Preparing data for analysis

### Type conversion
```{r}
# load dataset
students <- read.csv(f('students_with_dates.csv'), header=TRUE)
names(students)
```
```{r}
# Coerce Grades to character
students$Grades <- as.character(students$Grades) 

# Coerce Medu to factor
students$Medu <- as.factor(students$Medu)

# Coerce Fedu to factor
students$Fedu <- as.factor(students$Fedu)
```
### dates, lubridate package
these functions combine the letters y, m, d, h, m, s, which stand for year, month, day, hour, minute, and second, respectively. The order of the letters in the function should match the order of the date/time you are attempting to read in, although not all combinations are valid. 
```{r}
class(students$nurse_visit)
```
```{r}
# load the lubridate package
library(lubridate)

# convert nurse_visit to date format
students$nurse_visit <- ymd_hms(students$nurse_visit)
students$nurse_visit[1:3]
```
### stringr package

#### Trimming and padding strings
One common issue that comes up when cleaning data is the need to remove leading and/or trailing white space. The str_trim() function from stringr makes it easy to do this while leaving intact the part of the string that you actually want.
```{r}
# Load the stringr package
library(stringr)

# Trim all leading and trailing whitespace
str_trim(c("   Filip ", "Nick  ", " Jonathan"))
```
A similar issue is when you need to pad strings to make them a certain number of characters wide.
```{r}
# Pad these strings with leading zeros
str_pad(c("23485W", "8823453Q", "994Z"), width=9, side='left', pad=0)
```
#### Finding and replacing strings
The stringr package provides two functions that are very useful for finding and/or replacing patterns in strings: str_detect() and str_replace().
```{r}
# Detect all dates of nurse visits in 1997
# str_detect(students$nurse_visit, '1997')

# In the sex column, replace "F" with "Female" ...
students$sex <- str_replace(students$sex, 'F', 'Female')

# ... and "M" with "Male"
students$sex <- str_replace(students$sex, 'M', 'Male')
```
### Missing and special values

#### finding missing values
```{r}
df <- data.frame(A = c(1, NA, 8, NA),
                 B = c(3, NA, 88, 23),
                 C = c(2, 45, 1, 3))

# check for NAs
is.na(df)
```
```{r}
# are there any NAs?
any(is.na(df)) 
```
```{r}
#count number of NAs
sum(is.na(df))
```
```{r}
summary(df)
```
```{r}
#find rows with no missing values
complete.cases(df)
```
```{r}
# subset data, keeping only complete cases
df[complete.cases(df), ]
```
```{r}
# another way to remove rows with NAs
na.omit(df)
```

### Outliers and obvious errors

