# sales.csv case study
Lets load in the DF. We do not knwo anything about the factor variables so we set all character/string values to be processed just strings.
```{r}
source(file = 'funcs.R')
sales <- read.csv(file = f('sales.csv'), stringsAsFactors = FALSE)
```
After we have read the DF we inspect it via few commands:

### str()
### summary()
### dim()
### names()
### info() from funcs.R
```{r}
info(sales)
```
We see that the firts column "X" has no use, it counts the row index. We thus can omit it simply via:
```{r}
# substitute sales, not to contain the first column 'X'
sales2 <- sales[, -1]
head(sales2, n=3)
```
## Information not worth keeping

Many of the columns have information that's of no use to us. For example, the first four columns contain internal codes representing particular events. The last fifteen columns also aren't worth keeping; there are too many missing values to make them worthwhile.

An easy way to get rid of unnecessary columns is to create a vector containing the column indices you want to keep, then subset the data based on that vector using single bracket subsetting
```{r}
keep <- c(5 : (ncol(sales2)-15))
sales3 <- sales2[, keep]
```

## Separate columns
There are columns that contain multiple pieces of information which needs to be seprarated into two different columns. For example we have date and time in a single column:
```{r}
head(sales3$event_date_time, n=2)
```
### separate()
```{r}
library(tidyr)
# Split event_date_time: sales4
sales4 <- separate(data = sales3, col = event_date_time, into = c("event_dt", "event_time"), sep = " ")

# Split sales_ord_create_dttm: sales5
sales5 <- separate(sales4, sales_ord_create_dttm, c("ord_create_dt", "ord_create_time"), sep= " ")
```
Looks like that second call to separate() threw a warning. Lets inspect:
```{r}
subset(sales5[c(2516, 3863, 4082, 4183), ], select = c(event_dt, event_time, ord_create_dt, ord_create_time))
```
```{r}
# Define an issues vector
issues <- c(2516, 3863, 4082, 4183)

# Print values of sales_ord_create_dttm at these indices
sales3$sales_ord_create_dttm[issues]
```
The warning was just because of four missing values. 

## Identifying dates
Some of the columns in your dataset contain dates of different events. Right now, they are stored as character strings. That's fine if all you want to do is look up the date associated with an event, but if you want to do any comparisons or math with the dates, it's MUCH easier to store them as Date objects.

Luckily, all of the date columns in this dataset have the substring "dt" in their name, so you can use the str_detect() function of the stringr package to find the date columns. Then you can coerce them to Date objects using a function from the lubridate package.

### str_detect()
### ymd()

```{r}
# Load stringr
library(stringr)

# Find columns of sales5 containing "dt": date_cols
date_cols <- str_detect(names(sales5), "dt")

# Load lubridate
library(lubridate)

# Coerce date columns into Date objects
sales5[, date_cols] <- lapply(sales5[, date_cols], ymd)
```
Some warnings! That's probably because of more missing data, but again, it's good to check to be sure.
```{r}
# Create logical vectors indicating missing values (don't change)
missing <- lapply(sales5[, date_cols], is.na)

# Create a numerical vector that counts missing values: num_missing
num_missing <- sapply(missing, sum)

# Print num_missing
num_missing
```
As your last step, you'll use the tidyr function unite() to combine the venue_city and venue_state columns into one column with the two values separated by a comma and a space. For example, "PORTLAND" "MAINE" should become "PORTLAND, MAINE".

### unite()
```{r}
head(sales5[, c('venue_city', 'venue_state')], n=3)
```

```{r}
# Combine the venue_city and venue_state columns
sales6 <- unite(sales5, col="venue_city_state", c("venue_city", "venue_state"), sep=", ")

head(sales6[, 'venue_city_state'], n=3)
```

# Using readxl
The Massachusetts Bay Transportation Authority ("MBTA" or just "the T" for short) manages America's oldest subway, as well as Greater Boston's commuter rail, ferry, and bus systems.
```{r}
source('funcs.R')
library(readxl)
library(dplyr)

mbta <- read_excel(path = f('mbta.xlsx'), skip = 1)
```
It appears that the data are organized with observations stored as columns rather than as rows. You can fix that.

First, though, you can address the missing data. All of the NA values are stored in the All Modes by Qtr row. This row really belongs in a different data frame; it is a quarterly average of weekday MBTA ridership. Since this dataset tracks monthly average ridership, you'll remove that row.

Similarly, the 7th row (Pct Chg / Yr) and the 11th row (TOTAL) are not really observations as much as they are analysis. Go ahead and remove the 7th and 11th rows as well.
```{r}
mbta[c(1, 7, 11),]
```
```{r}
# Remove rows 1, 7, and 11 of mbta: mbta2
mbta2 <- mbta[c(-1, -7, -11),]

# Remove the first column of mbta2: mbta3
mbta3 <- mbta2[, -1]
```

## Observations are stored in columns
```{r}
mbta3[1,]
```

As is customary, you want to represent variables in columns rather than rows. The first step is to use the gather() function from the tidyr package, which will gather columns into key-value pairs. Mode column is allready a variable so this has to stay there.
```{r}
mbta4 <- gather(data = mbta3, key = 'month', value = 'thou_riders', -mode)
mbta4[1,]
```
## Type conversion
```{r}
# Coerce thou_riders to numeric
mbta4$thou_riders <- as.numeric(mbta4$thou_riders)
```
Now, you can finish the job you started earlier: getting variables into columns. Right now, variables are stored as "keys" in the mode column. You'll use the tidyr function spread() to make them into columns containing average weekday ridership for the given month and mode of transport.
```{r}
# Spread the contents of mbta4: mbta5
mbta5 <- spread(data = mbta4, key = mode, value = thou_riders)

mbta5[1,]
```
## Separating columns

Your dataset is already looking much better! Your boss saw what a great job you're doing and now wants you to do an analysis of the T's ridership during certain months across all years.

Your dataset has month names in it, so that analysis will be a piece of cake. There's only one small problem: if you want to look at ridership on the T during every January (for example), the month and year are together in the same column, which makes it a little tricky.

In this exercise, you'll separate the month column into distinct month and year columns to make life easier.
```{r}
mbta6 <- separate(data = mbta5, col = month, into = c('year', 'month'), sep = '-')
mbta6[1,]
```
```{r}
# View a summary of mbta6
summary(mbta6)

# Generate a histogram of Boat column
hist(mbta6$Boat)
```
Think for a minute about that Boat histogram. Every month, average weekday commuter boat ridership was on either side of four thousand. Then, one month it jumped to 40 thousand without warning?

Unless the Olympics were happening in Boston that month (they weren't), this value is certainly an error. You can assume that whoever was entering the data that month accidentally typed 40 instead of 4.
```{r}
# Find the row number of the incorrect value: i
i <- which(mbta6$Boat > 20)

# Replace the incorrect value with 4
mbta6$Boat[i] <- 4

# Generate a histogram of Boat column
hist(mbta6$Boat)
```
```{r}
mbta7 <- unite(mbta6, col = month, c('year', 'month'), sep = '')
mbta8 <- mbta7[, c('month', 'Boat', 'Trackless Trolley')]
mbta_boat <- gather(mbta8, key = mode, value = thou_riders, -month)
mbta_all <- gather(mbta7, key=mode, value = thou_riders, -month)
```


```{r}
library(ggplot2)
# Look at Boat and Trackless Trolley ridership over time
ggplot(mbta_boat, aes(x = month, y = thou_riders, col = mode)) +  geom_point() + 
  scale_x_discrete(name = "Month", breaks = c(200701, 200801, 200901, 201001, 201101)) + 
  scale_y_continuous(name = "Avg Weekday Ridership (thousands)")
```
```{r}
# Look at all T ridership over time (don't change)
ggplot(mbta_all, aes(x = month, y = thou_riders, col = mode)) + geom_point() + 
  scale_x_discrete(name = "Month", breaks = c(200701, 200801, 200901, 201001, 201101)) +  
  scale_y_continuous(name = "Avg Weekday Ridership (thousands)")
```


# multicolumn dataset

```{r}
library(gdata)

# Import the spreadsheet: att
att <- read.xls(f('attendance.xls'))
```
```{r}
names(att)
```
```{r}
# Create remove
remove <- c(3, 56:59)

# Create att2
att2 <- att[-remove,]
```
```{r}
# Create remove
remove <- c(3, 5, 7, 9, 11, 13, 15, 17)

# Create att3
att3 <- att2[, -remove]
```
In this data frame, columns 1, 6, and 7 represent attendance data for US elementary schools, columns 1, 8, and 9 represent data for secondary schools, and columns 1 through 5 represent data for all schools in the US.

Each of these should be stored as its own separate data frame, so you'll split them up here.
```{r}
# Subset just elementary schools: att_elem
att_elem <- att3[, c(1,6,7)]

# Subset just secondary schools: att_sec
att_sec <- att3[, c(1,8,9)]

# Subset all schools: att4
att4 <- att3[, 1:5] 
```
```{r}
names(att4)
```
```{r}
# Define cnames vector (don't change)
cnames <- c("state", "avg_attend_pct", "avg_hr_per_day", 
            "avg_day_per_yr", "avg_hr_per_yr")

# Assign column names of att4
colnames(att4) <- cnames

# Remove first two rows of att4: att5
att5 <- att4[-c(1,2),]

# View the names of att5
names(att5)
```
```{r}
# Remove all periods in state column
att5$state <- str_replace_all(att5$state, "\\.", "")

# Remove white space around state names
att5$state <- str_trim(att5$state, side='both')

# View the head of att5
head(att5)
```

