---
output:
  pdf_document: default
  html_document: default
---
# dplyr
```{r}
library(dplyr)
library(hflights)
```

## tbls

### as_tibble()
Convert data.frame to tibble.
```{r}
hflights <- as_tibble(hflights)

hflights
class(hflights)
```
## chaning labels
```{r}
unique(hflights$UniqueCarrier)
```
```{r}
lut <- c("AA" = "American", "AS" = "Alaska", "B6" = "JetBlue", "CO" = "Continental", 
         "DL" = "Delta", "OO" = "SkyWest", "UA" = "United", "US" = "US_Airways", 
         "WN" = "Southwest", "EV" = "Atlantic_Southeast", "F9" = "Frontier", 
         "FL" = "AirTran", "MQ" = "American_Eagle", "XE" = "ExpressJet", "YV" = "Mesa")

# Add the Carrier column to hflights
hflights$Carrier <- lut[hflights$UniqueCarrier]

unique(hflights$Carrier)
```
```{r}
glimpse(hflights$UniqueCarrier)
```
```{r}
glimpse(hflights$Carrier)
```
Change the labels in the CancellationCode column. This column lists reasons why a flight was cancelled using a non-informative alphabetical code.
```{r}
# The lookup table
lut <- c("A" = "carrier", "B" = "weather", "C" = "FFA", "D" = "security", "E" = "not cancelled")

# Add the Code column
hflights$Code <- lut[hflights$CancellationCode]

#glimpse it
unique(hflights[, c("CancellationCode", "Code")])
```
## 5 verbs

### select()
returns a subset of the columns,
```{r}
# Print out a tbl with the four columns of hflights related to delay
select(hflights, 'ActualElapsedTime', 'AirTime', 'ArrDelay', 'DepDelay')

# Print out the columns Origin up to Cancelled of hflights
select(hflights, 'Origin':'Cancelled')
```

#### Helper functions
dplyr comes with a set of helper functions that can help you select groups of variables inside a select() call:

- **starts_with("X")**: every name that starts with "X",
- **ends_with("X")**: every name that ends with "X",
- **contains("X")**: every name that contains "X",
- **matches("X")**: every name that matches "X", where "X" can be a regular expression,
- **num_range("x", 1:5)**: the variables named x01, x02, x03, x04 and x05,
- **one_of(x)**: every name that appears in x, which should be a character vector.

Pay attention here: When you refer to columns directly inside select(), you don't use quotes. If you use the helper functions, you do use quotes.

### filter()
return a subset of the rows, 
```{r}
# All flights that traveled 3000 miles or more
x1 <- filter(hflights, Distance >= 3000)

# All flights flown by JetBlue, Southwest, or Delta
x2 <- filter(hflights, UniqueCarrier %in% c("JetBlue", "Southwest", "Delta"))

# All flights where taxiing took longer than flying
x3 <- filter(hflights, TaxiIn + TaxiOut > AirTime)
```


### arrange()
that reorders the rows according to single or multiple variables,
```{r}
# Definition of dtc
dtc <- filter(hflights, Cancelled == 1, !is.na(DepDelay))

# Arrange dtc by departure delays
arrange(dtc, DepDelay)

# Arrange dtc so that cancellation reasons are grouped
arrange(dtc, CancellationCode)

# Arrange dtc according to carrier and departure delays
arrange(dtc, UniqueCarrier, DepDelay)
```
Reverse order of arrange
```{r}
# Arrange according to carrier and decreasing departure delays
arrange(hflights, UniqueCarrier, desc(DepDelay))
```


### mutate()
add columns from existing data.
```{r}
# Add the new variable ActualGroundTime to a copy of hflights and save the result as g1.
g1 <- mutate(hflights, ActualGroundTime = ActualElapsedTime - AirTime)
```


### summarize()
 reduces each group to a single row by calculating aggregate measures.
```{r}
# Print out a summary with variables min_dist and max_dist
summarize(hflights, min_dist = min(Distance), max_dist = max(Distance))

# Print out a summary with variable max_div
summarize(filter(hflights, Diverted == 1), max_div = max(Distance))
```

## Aggregate functions

- **min(x)** - minimum value of vector x.
- **max(x)** - maximum value of vector x.
- **mean(x)** - mean value of vector x.
- **median(x)** - median value of vector x.
- **quantile(x, p)** - pth quantile of vector x.
- **sd(x)** - standard deviation of vector x.
- **var(x)** - variance of vector x.
- **IQR(x)** - Inter Quartile Range (IQR) of vector x.
- **diff(range(x))** - total range of vector x.

## dplyr aggregate functions

- **first(x)** - The first element of vector x.
- **last(x)** - The last element of vector x.
- **nth(x, n)** - The nth element of vector x.
- **n()** - The number of rows in the data.frame or group of observations that summarize() describes.
- **n_distinct(x)** - The number of unique values in vector x.

```{r}
# Generate summarizing statistics for hflights
summarize(hflights,
          n_obs = n(),
          n_carrier = n_distinct(UniqueCarrier),
          n_dest = n_distinct(Dest))

# All American Airline flights
aa <- filter(hflights, UniqueCarrier == "American")

# Generate summarizing statistics for aa 
summarize(aa,
          n_flights = n(),
          n_canc = sum(Cancelled == 1),
          avg_delay = mean(ArrDelay, na.rm=TRUE))
```

## pipe operator %>%
    Take the hflights data set and then ...
    Add a variable named diff that is the result of subtracting TaxiIn from TaxiOut, and then ...
    Pick all of the rows whose diff value does not equal NA, and then ...
    Summarize the data set with a value named avg that is the mean diff value.
```{r}
hflights %>%
  mutate(diff = TaxiOut - TaxiIn) %>% 
  filter(!is.na(diff)) %>% 
  summarize(avg = mean(diff))
```
Count the number of ovrnight flights
```{r}
hflights %>% 
  filter(!is.na(DepTime), !is.na(ArrTime), DepTime > ArrTime) %>% 
  summarize(num = n())
```

## group_by()
### rank()
```{r}
# Ordered overview of average arrival delays per carrier
hflights %>%
  filter(!is.na(ArrDelay) & ArrDelay > 0) %>%
  group_by(UniqueCarrier) %>%
  summarize(avg = mean(ArrDelay)) %>%
  mutate(rank = rank(avg)) %>%
  arrange(rank)
```
## dplyr through mySQL database
```{r}
# library(RMySQL)
# library(SQ)
# # Set up a connection to the mysql database
# my_db <- src_mysql(dbname = "dplyr", 
#                    host = "courses.csrrinzqubik.us-east-1.rds.amazonaws.com", 
#                    port = 3306, 
#                    user = "student",
#                    password = "datacamp")
# 
# # Reference a table within that source: nycflights
# nycflights <- tbl(my_db, "dplyr")
# 
# # glimpse at nycflights
# glimpse(nycflights)
# 
# # Ordered, grouped summary of nycflights
# nycflights %>%
#   group_by(carrier) %>%
#   summarize(n_flights = n(),
#             avg_delay = mean(arr_delay, na.rm = TRUE)) %>%
#   arrange(avg_delay)
```


