---
output:
  pdf_document: default
  html_document: default
---
# Importing data in R

## Flat files (csv)
```{r}
# Function to enter file paths fast:
path <- '/home/ilves/Documents/R_Rstudio/R/data_files/'

f <- function(x) {
  # x : string, filename to read
  path <- '/home/ilves/Documents/R_Rstudio/R/data_files/'
  paste(path, x, sep='')
}
```


```{r}

# univ_path <- file.path("~", "")

pools <- read.csv(f('swimming_pools.csv'), stringsAsFactors = FALSE) 
# stringsAsFactors = TRUE converts values to categorical variables

str(pools)
```
### tab-delimited files
```{r}
hotdogs <- read.delim(f('hotdogs.txt'), header=FALSE)

summary(hotdogs)
```

### excotic sep format
read any tabular file into df
```{r}
hotdogs <- read.table(f('hotdogs.txt'), 
                      sep = '\t', 
                      col.names = c("type", "calories", "sodium"))
head(hotdogs)
```
```{r}
# Select the hot dog with the least calories: lily
lily <- hotdogs[which.min(hotdogs$calories), ]

# which. returns the index of inquired variable

# Select the observation with the most sodium: tom
tom <- hotdogs[which.max(hotdogs$sodium), ]
cat(paste(lily, tom), sep = '\n\n')
```

### Column classes
If a column is set to "NULL" in the colClasses vector, this column will be skipped and will not be loaded into the data frame.
```{r}
hotdogs2 <- read.delim(f("hotdogs.txt"), header = FALSE, 
                       col.names = c("type", "calories", "sodium"),
                       colClasses = c("factor", "NULL", "numeric"))


# Display structure of hotdogs2
str(hotdogs2)
```

# readr package()
```{r}
library(readr)

read_csv(f('potatoes.csv'))
```
```{r}
# Column names
properties <- c("area", "temp", "size", "storage", "method",
                "texture", "flavor", "moistness")

# Import potatoes.txt: potatoes
potatoes <- read_tsv(f('potatoes.txt'), col_names=properties)

# Call head() on potatoes
head(potatoes)
```
read_delim() is the main readr function
```{r}
# Column names
properties <- c("area", "temp", "size", "storage", "method",
                "texture", "flavor", "moistness")

# Import potatoes.txt using read_delim(): potatoes
potatoes <- read_delim(f('potatoes.txt'), delim='\t', col_names=properties)

# Print out potatoes
potatoes
```
## skip, n_max
Through skip and n_max you can control which part of your flat file you're actually importing into R.

* skip specifies the number of lines you're ignoring in the flat file before actually starting to import data.
* n_max specifies the number of lines you're actually importing.

**Watch out**: Once you skip some lines, you also skip the first line that can contain column names!
```{r}
# Column names
properties <- c("area", "temp", "size", "storage", "method",
                "texture", "flavor", "moistness")

# Import 5 observations from potatoes.txt: potatoes_fragment
potatoes_fragment <- read_tsv(f("potatoes.txt"), skip = 6, n_max = 5, col_names = properties)
```
# data.table package
* fast

## fread()
**drop** and **select** arguments, to drop or select variables of interest by position or by column name
```{r}
library(data.table)
# Import columns 6 and 8 of potatoes.csv: potatoes
potatoes <- fread(f('potatoes.csv'), select=c(6,8))

# Plot texture (x) and moistness (y) of potatoes
plot(x=potatoes$texture, y=potatoes$moistness)
```
# readxl

## List the sheets of an Excel file
```{r}
# Load the readxl package
library(readxl)

# Print the names of all worksheets
excel_sheets(f('urbanpop.xlsx'))
```

## Import an Excel sheet
```{r}
# Read the sheets, one by one
pop_1 <- read_excel(f("urbanpop.xlsx"), sheet = 1)
pop_2 <- read_excel(f("urbanpop.xlsx"), sheet = 2)
pop_3 <- read_excel(f('urbanpop.xlsx'), sheet = 3)

# Put pop_1, pop_2 and pop_3 in a list: pop_list
pop_list <- list(pop_1, pop_2, pop_3)

# Display the structure of pop_list
str(pop_list)[1:5]
```
## reading a workbook
Example code:
```{r}
# my_workbook <- lapply(excel_sheets("data.xlsx"),
#                       read_excel,
#                       path = "data.xlsx")
```
The read_excel() function is called multiple times on the "data.xlsx" file and each sheet is loaded in one after the other. The result is a list of data frames, each data frame representing one of the sheets in data.xlsx.
```{r}
# Read all Excel sheets with lapply(): pop_list
pop_list <- lapply(excel_sheets(f('urbanpop.xlsx')),
read_excel, path=f('urbanpop.xlsx'))

# Display the structure of pop_list
str(pop_list)
```
## remove NA values from DFs
```{r}
# na.omit(DF)
```
# XLConnect
```{r}
library(XLConnect)

# Build connection to urbanpop.xlsx
my_book <- loadWorkbook(f("urbanpop.xlsx"))

# List the sheets in my_book
getSheets(my_book)
```
## createSheet()
```{r}
my_book <- loadWorkbook(f('urbanpop.xlsx'))
getSheets(my_book)
```
```{r}
createSheet(my_book, 'data_summary')
getSheets(my_book)
```

## writeWorksheet()
```{r}
# Create data frame: summ
sheets <- getSheets(my_book)[1:3]
dims <- sapply(sheets, function(x) dim(readWorksheet(my_book, sheet = x)), USE.NAMES = FALSE)
summ <- data.frame(sheets = sheets,
                   nrows = dims[1, ],
                   ncols = dims[2, ])

# Add data in summ to "data_summary" sheet
writeWorksheet(my_book, data=summ, sheet='data_summary')
```

## saveWorkbook()
```{r}
saveWorkbook(my_book, f('summary.xlsx'))
```

## reneameSheet()
```{r}
my_book <- loadWorkbook(f('summary.xlsx'))
getSheets(my_book)
```
```{r}
# Rename "data_summary" sheet to "summary"
renameSheet(my_book, sheet='data_summary', newName='summary')

# Print out sheets of my_book
getSheets(my_book)
```

## removeSheet()
```{r}
# Remove the fourth sheet
removeSheet(my_book, sheet='summary')

getSheets(my_book)
```

