---
output:
  pdf_document: default
  html_document: default
---
```{r}
library(gapminder)
library(dplyr)
```
```{r}
head(gapminder)
```
# Verbs
Every time verbs are used we produce a **%>%** pipe syntax. It says: 'What ever is before it and feed it into next step.'
For example:

## filter()
The filter verb extracts particular observations based on a condition.
```{r}
gapminder %>%
  filter(year == 2007, country == 'United States')
```

## arrange()
Changing order of a variable in the DF. Ascending:
```{r}
gapminder %>%
  arrange(gdpPercap)
```
Decending:
```{r}
gapminder %>%
  arrange(desc(gdpPercap))
```
Filtering for multiple things:
```{r}
gapminder %>%
  filter(year == 2007) %>%
  arrange(desc(gdpPercap))
```

## mutate()
Change or add new variable (column) to a DF. Change a variable:
```{r}
gapminder %>%
  mutate(pop = pop / 1000000)
```
Add a variable:
```{r}
gapminder %>%
  mutate(gdp = gdpPercap * pop)
```
# Visualization
```{r}
library(ggplot2)

# Create gapminder_1952
gapminder_1952 <- gapminder %>%
  filter(year == 1952) %>%
  mutate(pop = pop * 1000000)
```
```{r}
ggplot(gapminder_1952, aes(x=pop, y=gdpPercap)) + geom_point() + scale_x_log10() + scale_y_log10()
```
## Color of the aesthetic
```{r}
ggplot(gapminder_1952, 
       aes(x=gdpPercap, y=lifeExp, color=continent, size=pop)) + 
  geom_point() + 
  scale_x_log10()
```
## Faceting
Subplotting. 
```{r}
ggplot(gapminder_1952, aes(x=pop, y=lifeExp)) + 
  geom_point() + 
  scale_x_log10() + 
  facet_wrap(~ continent)
```

## summarize()
```{r}
gapminder %>%
  filter(year == 2007) %>%
  summarize(maxpop = max(pop), totalpop = sum(pop))
```

## group_by()

```{r}
gapminder %>%
  group_by(year) %>%
  summarize(medianLifeExp = median(lifeExp), maxGdpPercap = max(gdpPercap))
```
```{r}
gapminder %>%
  filter(year == 1957) %>%
  group_by(continent) %>%
  summarize(medianLifeExp = median(lifeExp), maxGdpPercap = max(gdpPercap))
```

# Visualizing summarized data

## line plots
```{r}
by_year <- gapminder %>%
  group_by(year) %>%
  summarize(medianGdpPercap = median(gdpPercap))

# Create a line plot showing the change in medianGdpPercap over time
ggplot(by_year, aes(x=year, y=medianGdpPercap)) + geom_line() + expand_limits(y = 0)
```
```{r}
by_year_continent <- gapminder %>%
  group_by(year, continent) %>%
  summarize(medianGdpPercap = median(gdpPercap))

# Create a line plot showing the change in medianGdpPercap by continent over time
ggplot(by_year_continent, aes(x=year, y=medianGdpPercap, color=continent)) + geom_line() + expand_limits(y = 0)
```

## bar plots
```{r}
by_continent <- gapminder %>%
  filter(year == 1952) %>%
  group_by(continent) %>%
  summarize(medianGdpPercap = median(gdpPercap))

# Create a bar plot showing medianGdp by continent
ggplot(by_continent, aes(x=continent, y=medianGdpPercap)) + geom_col()
```

## Histograms
```{r}
gapminder_1952 <- gapminder %>%
  filter(year == 1952)

# Create a histogram of population (pop), with x on a log scale
ggplot(gapminder_1952, aes(x=pop)) + geom_histogram() + scale_x_log10()
```

## box plots
```{r}
gapminder_1952 <- gapminder %>%
  filter(year == 1952)

# Create a boxplot comparing gdpPercap among continents
ggplot(gapminder_1952, aes(x=continent, y=gdpPercap)) + geom_boxplot() + scale_y_log10()
```

