---
output:
  pdf_document: default
  html_document: default
---
## Variable assignment
```{r}
x <- 42
x
```
## Basic data types
* Decimal values like **4.5** are **numerics**
* Natural numbers like **4** are **integers**
* Boolean values **TRUE/FALSE** are **logical**
* **Text** values are **characters**

## Data type checking
```{r}
my_logical <- TRUE
class(my_logical)
```
## Vectors
Vectors are one-dimension arrays that can hold numeric data, character data, or logical data. 
* In R, you create a vector with the combine function **c()**
```{r}
numeric_vector <- c(1, 10, 49)
character_vector <- c("a", "b", "c")
boolean_vector <- c(TRUE, FALSE, TRUE)
numeric_vector
character_vector
boolean_vector
```
## Naming a vector
some_vector <- c("John Doe", "poker player")
names(some_vector) <- c("Name", "Profession")
```{r}
some_vector <- c("John Doe", "poker player")
names(some_vector) <- c("Name", "Profession")
some_vector
```
## Selecting elements from a vector
* Selecting a single value: c[2]
* Selecting mutiple values c[c(2,3)]
```{r}
some_vector[1]
character_vector[c(1,3)]
```
## Selection by comparison
Selecting some values of a vector with TRUE logic:
```{r}
poker_vector <- c(140, -50, 20, -120, 240)
days_vector <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
names(poker_vector) <- days_vector
selection_vector <- poker_vector > 0

poker_winning_days <- poker_vector[selection_vector]
poker_winning_days
```
## Matrix
In R, a **matrix** is a collection of elements of **the same data type** (numeric, character, or logical) arranged into a fixed number of rows and columns. Since you are only working with rows and columns, a matrix is called two-dimensional. 
```{r}
matrix(1:9, byrow=TRUE, nrow=3)
```
## Naming the matrix rows and columns
```{r}
# Box office Star Wars (in millions!)
new_hope <- c(460.998, 314.4)
empire_strikes <- c(290.475, 247.900)
return_jedi <- c(309.306, 165.8)

# Construct matrix
star_wars_matrix <- matrix(c(new_hope, empire_strikes, return_jedi), nrow = 3, byrow = TRUE)

# Vectors region and titles, used for naming
region <- c("US", "non-US")
titles <- c("A New Hope", "The Empire Strikes Back", "Return of the Jedi")

# Name the columns with region
colnames(star_wars_matrix) <- region

# Name the rows with titles
rownames(star_wars_matrix) <- titles

# Print out star_wars_matrix
star_wars_matrix
```
## Adding a column to a matrix
```{r}
# Construct star_wars_matrix
box_office <- c(460.998, 314.4, 290.475, 247.900, 309.306, 165.8)
star_wars_matrix <- matrix(box_office, nrow = 3, byrow = TRUE,
                           dimnames = list(c("A New Hope", "The Empire Strikes Back", "Return of the Jedi"), 
                                           c("US", "non-US")))

# The worldwide box office figures
worldwide_vector <- rowSums(star_wars_matrix)

# Bind the new variable worldwide_vector as a column to star_wars_matrix
all_wars_matrix <- cbind(star_wars_matrix, worldwide_vector)
all_wars_matrix
```
## Adding a row to matrix
rbind(matrix1, matrix2)

## Factors
The term factor refers to a statistical data type used to store **categorical variables**. The difference between a categorical variable and a continuous variable is that a categorical variable can belong to a limited number of categories. A continuous variable, on the other hand, can correspond to an infinite number of values.
```{r}
# Sex vector
sex_vector <- c("Male", "Female", "Female", "Male", "Male")

# Convert sex_vector to a factor
factor_sex_vector <- factor(sex_vector)

# Print out factor_sex_vector
factor_sex_vector
```
### Nominal categorical variable
A nominal variable is a categorical variable without an implied order. This means that it is impossible to say that 'one is worth more than the other'. For example, think of the categorical variable animals_vector with the categories "Elephant", "Giraffe", "Donkey" and "Horse". Here, it is impossible to say that one stands above or below the other.
```{r}
# Animals
animals_vector <- c("Elephant", "Giraffe", "Donkey", "Horse")
factor_animals_vector <- factor(animals_vector)
factor_animals_vector
```

### Ordinal categorical variable
In contrast, ordinal variables do have a natural ordering. Consider for example the categorical variable temperature_vector with the categories: "Low", "Medium" and "High". Here it is obvious that "Medium" stands above "Low", and "High" stands above "Medium".
```{r}
# Temperature
temperature_vector <- c("High", "Low", "High","Low", "Medium")
factor_temperature_vector <- factor(temperature_vector, order = TRUE, levels = c("Low", "Medium", "High"))
factor_temperature_vector
```
## Specifing the levels
he order with which you assign the levels is important. If you type levels(factor_survey_vector), you'll see that it outputs [1] "F" "M". If you don't specify the levels of the factor when creating the vector, R will automatically assign them alphabetically. 
```{r}
# Code to build factor_survey_vector
survey_vector <- c("M", "F", "F", "M", "M")
factor_survey_vector <- factor(survey_vector)

# Specify the levels of factor_survey_vector
levels(factor_survey_vector) <- c('Female', 'Male')

factor_survey_vector
```
## summary() function
```{r}
summary(factor_survey_vector)
```
```{r}
summary(survey_vector)
```
## Ordered factors example
```{r}
# Create speed_vector
speed_vector <- c("medium", "slow", "slow", "medium", "fast")

# Convert speed_vector to ordered factor vector
factor_speed_vector <- factor(speed_vector, 
                              ordered=TRUE, 
                              levels=c('slow', 'medium', 'fast'))
factor_speed_vector
```
# DataFrame
## Creating a DF
```{r}
# Definition of vectors
name <- c("Mercury", "Venus", "Earth", "Mars", "Jupiter", "Saturn", "Uranus", "Neptune")
type <- c("Terrestrial planet", "Terrestrial planet", "Terrestrial planet", 
          "Terrestrial planet", "Gas giant", "Gas giant", "Gas giant", "Gas giant")
diameter <- c(0.382, 0.949, 1, 0.532, 11.209, 9.449, 4.007, 3.883)
rotation <- c(58.64, -243.02, 1, 1.03, 0.41, 0.43, -0.72, 0.67)
rings <- c(FALSE, FALSE, FALSE, FALSE, TRUE, TRUE, TRUE, TRUE)

# Create a data frame from the vectors
planets_df <- data.frame(name, type, diameter, rotation, rings)
head(planets_df)
```
## Structure
```{r}
str(planets_df)
```
## Select a column
```{r}
planets_df[,1]
```
```{r}
planets_df[, 'name']
```
```{r}
planets_df['name']
```
```{r}
planets_df$name
```
## subset()
```{r}
# Select planets with diameter < 1
subset(planets_df, diameter < 1)
```
## Sorting
```{r}
a = c(100, 10, 1, 1000)
a[order(a)]
```
```{r}
# Use order() to create positions
positions <-  order(planets_df$diameter)

# Use positions to sort planets_df
planets_df[positions,]
```
# Lists
A list in R allows you to gather a variety of objects under one name (that is, the name of the list) in an ordered way. These objects can be matrices, vectors, data frames, even other lists, etc.
## Named list
```{r}
# Vector with numerics from 1 up to 10
my_vector <- 1:10 

# Matrix with numerics from 1 up to 9
my_matrix <- matrix(1:9, ncol = 3)

# First 10 elements of the built-in data frame mtcars
my_df <- planets_df[1:4,]

# Adapt list() call to give the components names
my_list <- list(my_vector, my_matrix, my_df)
names(my_list) <- c('vec', 'mat', 'df')
# Print out my_list
print(my_list)
```


